Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-01 bmc_template/bmc_article.tex
temp pushing

>>>>

\section{Discussion}

HMMs are a widely used and flexible tool for modeling animal movement
behaviour; we need to work harder to make sure they are both appropriately
complex and biologically interpretable. 
With the increasing volumes of movement data available,
ecologists who naively use traditional homogeneous HMMs
and standard information-theoretic criteria
to estimate the number of behavioural states will generally
overfit their data, in the sense of ``discovering'' large
number of states that are difficult to interpret biologically.

On a broad spectrum, it really depends on what kind of question 
that is being answered. On one side of the spectrum, if the goal is 
to identify states, it might be sufficient to use a 
simple/traditional HMM model and pre-specify the number of states and, post hoc, 
match Viterbi-based states estimates with environmental variation 
\cite{fryxell_multiple_2008}. On the other side of the 
spectrum, if the goal of interest is to make predictions (out of sample), 
it might be better to fit a covariate-dependent model so that we can explicitly
model the switching process. In that case, fitting a covariate-dependent model is
better for out of sample prediction because Viterbi can only estimate state 
occupancy if observed movements are available (within sample preditions). 
Finally, if we want to estimate the number of states, BIC is not necessarily 
good for estimation of number of states \cite{biernacki2000assessing}, 
but it can be useful as an approximate upper limit estimate.

Incorporating temporal heterogenity in animal movement is one step in
the right direction, but much remains to be done. 
Our model neglects other predictors, such as habitat type or
location with respect to environmental features such as roads,
that can potentially improve goodness of fit and predictions and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward using existing
frameworks, including all possible biological complexities in 
a HMM with state-dependent transitions may rapidly become intractable
in terms of both computational time and complexity of choosing
among possible reduced models and numbers of states.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest.

\section{Conclusion}

We have presented a relatively simple but little-used extension
(time-dependent transitions) that partly resolves the problem.
Time-dependent transitions appear to offer a simple way to (1) reduce
the selected number of states closer to a biologically interpretable
level; (2) capture observed diurnal and autocorrelation patterns in a
predictive model; (3) improve overall model fit (i.e., lower BIC) and
reduce the level of complexity (number of parameters) of the most
parsimonious models. Simple simulations where the true number of
states is known, and transitions among states vary over time, confirm
that using BIC with homogeneous HMMs overestimates the number of
behavioural states, while time-dependent HMMs correctly estimate
the number.
